Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-23 doc/asplos13/memory.tex
First crack at normalizing memory after the redux

Applications written for Grappa utilize two forms of memory: local and
global.

Local memory is local to a single core in the system.  Accesses occur
through conventional pointers.  The compiler emits an access and the
memory is manipulated directly.  Applications use local accesses for a
number of things in Grappa: the stack associated with a task, accesses
to localized global memory in caches (see below), and accesses to
debugging infrastructure that is local to each system node.  Local
pointers cannot access memory on other cores, and are valid only on
their home core.

@@

\paragraph{Global memory allocation}
Grappa provides two methods for {\emph storing} data in the global memory. The
first is a distributed heap striped across all the machines in the
system. The \texttt{global\_malloc} and \texttt{global\_free} calls
are used to allocate and deallocate memory in the global heap; on
allocation, a global pointer is returned. Grappa also allows any local
data on a core's stacks or heap to be exported to the global address
space to be made accessible to other cores across the system.

@@

\paragraph{Global memory access} There are two approaches to {\emph
accessing} global memory. When the programmer expects a computation on
shared data to have spatial locality to exploit, {\em cache} operations
may be used. When there is no locality to exploit, {\em delegate}
operations are used.

All global addresses in Grappa are 64-bit values. Each byte of global
memory is associated with a {\em home core}; this core is responsible
for allocating and modifying that memory. Since today's commodity
processors support a 48-bit virtual address space \cite{AMD64}, Grappa
uses the spare bits to distinguish which address space a global address
belongs to, and to represent the home core of a global address.

Global memory in Grappa is exposed through two distinct, non-overlapping
address spaces; \textbf{2D global addresses}, optimized for global
access to node-local data, and \textbf{linear global addresses},
optimized for high aggregate random access bandwidth to low locality
data.

@@

\paragraph{2D global addresses} The first global address space is the {\em
two-dimensional} address space. This uses a traditional PGAS addressing model,
where each address is a tuple of a rank in the job (or global process ID) and
an address in that process. The lower 48 bits of the address hold a virtual
address in the process. The top bit is set to indicates that the reference is
a 2D address. This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^{15}$ endpoints.

Any node-local data can be made accessible by other nodes in the system
by wrapping the address and node ID into a 2D global address. This
address can then be accessed with a delegate and can also be cached by
other nodes. At the destination the address is converted into a
canonical x86 address by replacing the upper bits with the sign-extended
upper bit of the virtual address. 2D addresses may refer to memory
allocated from a single processes' heap or from a task's stack.
Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can
refer to other cores' memory.

\paragraph{Linear global addresses} The second global address space is
the {\em linear} address space, so named because nearby addresses can be
treated as adjacent even if they are stored on different nodes.
Addresses in the linear address space are partitioned across the cluster
in a block-cyclic fashion. Choosing the block size involves trading off
sequential bandwidth against aggregate random access bandwidth. Smaller
block sizes help spread data across all the memory controllers in the
cluster, but larger block sizes allow the locality-optimized memory
controllers to provide increased sequential bandwidth. The block size,
which is configurable, is typically set to 64 bytes, or the size of a
single hardware cache line, in order to exploit spatial locality when
available.

All linear addresses refer to data allocated from a global heap. The
heap metadata is stored on a single node. Currently all heap operations
serialize through this node; while this has been sufficient for our
benchmarks, in the future Grappa will provide parallel performance
through combining~\cite{MAMA,flatcombining}.

\TODO{Add a figure that depicts both forms of global addresses.}

All global memory operations are implemented with active messages.
Requesters construct a descriptor holding the current task ID and space
for a response, issue their request active message to the node given by
the request address, and suspend themselves. When the request arrives at
the address' home node, the operation is performed and a response active
message is sent. When the response arrives at the requesting node, the
descriptor is filled in with the results of the operation and the
requesting task is placed back on the ready queue.

@@

Under the hood, Grappa performs the mechanics of gathering chunks of
data from multiple system nodes and presenting a conventional appearing
linear block of memory as a local pointer into a cache. The strategy
employed is to issue all the constituent requests of a cache access
request and then yield until all responses have occurred.  Currently,
Grappa caches are \emph{not} coherent, requiring the programmer to
maintain consistent access to data.  Future work will develop a software
directory based coherence scheme to simplify consistent access to global
data.

@@

\paragraph{Example}

@@

Figure~\ref{fig:delegate-cache} shows an example of delegate and cache
operations. In step 1, a core allocates an array in the global
heap. It then spawns two tasks on remote cores to increment elements
of the array.

>>>>

Applications written for Grappa utilize two forms of memory: local and
global.  Local memory is local to a single core in the system.  Accesses occur
through conventional pointers.  The compiler emits an access and the
memory is manipulated directly.  Applications use local accesses for a
number of things in Grappa: the stack associated with a task, accesses
to localized global memory in caches (see below), and accesses to
debugging infrastructure that is local to each system node.  Local
pointers cannot access memory on other cores, and are valid only on
their home core.

@@

\paragraph{Global memory addressing} Grappa provides two methods for
storing data in the global memory. The first is a distributed heap
striped across all the machines in the system in a block cyclic fashion.
The \texttt{global\_malloc} and \texttt{global\_free} calls are used to
allocate and deallocate memory in the global heap.  Addresses to memory
in the global heap use \textbf{linear addresses}.  Choosing the block
size involves trading off sequential bandwidth against aggregate random
access bandwidth. Smaller block sizes help spread data across all the
memory controllers in the cluster, but larger block sizes allow the
locality-optimized memory controllers to provide increased sequential
bandwidth. The block size, which is configurable, is typically set to 64
bytes, or the size of a single hardware cache line, in order to exploit
spatial locality when available. The heap metadata is stored on a single
node. Currently all heap operations serialize through this node; while
this has been sufficient for our benchmarks, in the future Grappa will
provide parallel performance through combining~\cite{MAMA,flatcombining}.

Grappa also allows any local data on a core's stacks or heap to be
exported to the global address space to be made accessible to other
cores across the system. Addresses to global memory allocated in this
way use \textbf{2D global addresses}.  This uses a traditional PGAS
addressing model, where each address is a tuple of a rank in the job (or
global process ID) and an address in that process. The lower 48 bits of
the address hold a virtual address in the process. The top bit is set to
indicates that the reference is a 2D address (as opposed to linear
address). This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^{15}$ endpoints. Any node-local data can be made
accessible by other nodes in the system by wrapping the address and node
ID into a 2D global address. This address can then be accessed with a
delegate and can also be cached by other nodes. At the destination the
address is converted into a canonical x86 address by replacing the upper
bits with the sign-extended upper bit of the virtual address. 2D
addresses may refer to memory allocated from a single processes' heap or
from a task's stack. Figure~\ref{fig:memory-structure} shows how 2D and
linear addresses can refer to other cores' memory.

@@

@@

\paragraph{Global memory access} There are two general approaches
Grappa applications use to {\emph access} global memory. When the
programmer expects a computation on shared data to have spatial locality
to exploit, {\em cache} operations may be used. When there is no
locality to exploit, {\em delegate} operations are used.

@@

Under the hood, Grappa performs the mechanics of gathering chunks of
data from multiple system nodes and presenting a conventional appearing
linear block of memory as a local pointer into a cache. The strategy
employed is to issue all the constituent requests of a cache access
request  (as Active Messages) and then yield until all responses have
occurred.  Currently, Grappa caches are \emph{not} coherent, requiring
the programmer to maintain consistent access to data.  Future work will
develop a software directory based coherence scheme to simplify
consistent access to global data.

@@

@@

\paragraph{Example} Figure~\ref{fig:delegate-cache} shows an example of
delegate and cache operations. In step 1, a core allocates an array in
the global heap. It then spawns two tasks on remote cores to increment
elements of the array.
